School closures and consolidations: the active travel and emission implications of reduced accessibility

### 2024-05-09 paper/Manuscript.qmd
Version submitted to journal

---
title: "School closures and consolidations: the active travel and emission implications of reduced accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
linenumbers: true
numbersections: true
format:
  nature-pdf: default
editor: source
execute: 
  freeze: auto
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
    attributes:
      corresponding: true
  - name: "Christopher D. Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliations:
      - id: 2
        address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"
  - name: "Antonio Paez"
    email: "paezha@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - name: "Moataz Mohamed"
    email: "mmohame@mcmaster.ca "
    affiliations:
      - id: 3
        address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
abstract: | 
  | Reducing emissions and increasing active travel are often top priorities for many communities. These priorities, though, often are confronted with a pressure to reduce the fiscal burden of public services, including schools. Public schools have for some time now been pressured to do more with less, and governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity. The objective of this study is to quantify the travel burdens of school consolidation/closure policy. We do this through the case of Hamilton, a mid-sized city in Ontario, Canada. Between the years 2011 and 2015, 7% of the elementary schools closed in the city. For this research, we study how these closures changed the accessibility landscape for the elementary school aged population, and the active travel and emission implications. In terms of methods, we use spatial availability, a singly-constrained multimodal measure of spatial accessibility. Thanks to its proportional allocation mechanism, spatial availability allows for an interpretable comparison between 'before' and 'after' the implementation of school closure/consolidation policy. Analysis is conducted at the parcel-level with all school seats being eligible and travel behaviour based on observed home-to-school patterns for motorized and non-motorized modes. We report estimates of system-wide impacts that result from the closures, such as potential growth in GHG emission and reduction of walking minutes. Overall, spatial availability decreased in the city. Though the school consolidation policy may have resulted in cost-savings in operation and maintenance of facilities, students, families, and communities pay the price in reduced opportunities for physical activity and greater pollution.
keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
---

@@

```{r, include=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggspatial) # Spatial Data Framework for ggplot2
library(here) # A Simpler Way to Find Your Files
library(patchwork) # The Composer of Plots
library(scales) # Scale Functions for Visualization
library(sf) # Simple Features for R
library(stplanr) # Sustainable Transport Planning
library(tidyverse) # Easily Install and Load the 'Tidyverse'
library(tmap) # Thematic Maps
library(units) # Measurement Units for R Vectors
library(miscTools) # Miscellaneous Tools and Utilities
library(huxtable)
```

@@

```{r descriptive-catchment-schools-plot2, eval=FALSE}
tmap_save(catchment_schools_plot, paste0(here::here(),"/Figures/catchment_schools_plot.png"), height=7, width=12, dpi=600)
```

@@

```{r student-popabs-lowincperc-poprate-plot2, eval=FALSE}
tmap_save(stud_pop_rate_LIM_plot, paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"), height=10, width=12, dpi=600)
```

@@

tmap_save(OD_flow_plot, paste0(here::here(),"/Figures/OD_flow_plot.png"), height=7, width=12, dpi=600)
```

@@

```{r ploting-empirical-and-theo-curve-TLD-tts, eval=FALSE}
TLD_empirical_wtheo_curve_plot <- ggplot() + 
  geom_histogram(data = od_trips_perc_tt,
                 aes(x = x, 
                     y = ..density..),
                 colour = "grey70",
                 fill = "grey70",
                 boundary = 0,
                 binwidth = 3) +
  geom_vline(data = q_1, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_2, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_3, mapping = aes(xintercept = x), colour = "grey60") +
  geom_line(data = tld_theoretical, aes(x = t, y = f), colour = "blue", size=0.8, alpha=0.6) + 
  xlab("Estimated travel time associated with TTS OD flows (minutes)") +
  ylab("Density") +
  scale_x_continuous(breaks = seq(0,45,10), expand=c(0,0))+
  scale_y_continuous(breaks = seq(0,0.12,0.04), limits = c(0,0.12))+
  facet_grid(~mode~year)+
  geom_text(data = q_1, 
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  geom_text(data = q_2,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  geom_text(data = q_3,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  theme_classic() +
  theme(axis.title = element_text(size = 11),
        panel.border = element_rect(colour = "grey80", fill=NA),
        strip.text = element_text(size=11))
TLD_empirical_wtheo_curve_plot
```

```{r ploting-empirical-and-theo-curve-TLD-tts2, eval=FALSE}
ggsave(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"), height=7, width=10, dpi=600)
```
```{r}
ALL_SCHOOLS_Elem_formapping2 <- ALL_SCHOOLS_Elem_formapping |> mutate(Mode = "motorized")
ALL_SCHOOLS_Elem_formapping3 <- ALL_SCHOOLS_Elem_formapping |> mutate(Mode = "non-motorized")
ALL_SCHOOLS_Elem_formapping2 <- rbind(ALL_SCHOOLS_Elem_formapping2, ALL_SCHOOLS_Elem_formapping3)
rm(ALL_SCHOOLS_Elem_formapping3)
```

@@

Vim_vvim_20112016_plot <- tmap_arrange(Vim_20112016_plot, vvim_20112016_plot, nrow=2)
tmap_save(Vim_vvim_20112016_plot, paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"), height=12, width=12, dpi=600)
```

@@

Vim_20112016_CHANGE_plot <-  
  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
  tm_polygons(col = "c_V_im", border.alpha = 0, border.col = "grey20", lwd=0.2,
              palette = "RdBu", #midpoint=0,
              # style="quantile", n=4,
              breaks = c(-20, -4.1, -0.2, 0, 20),
              labels = c("[-208 to -4.1)", "[-4.1 to -0.2)", "[-0.2 to 0)","[0 to 283.6]"),
              title = expression("Change in "*V[i]^m),
              textNA = "No V in 2016 or 2011",
              colorNA = "white")+
  tm_facets(by = c("mode"), nrow=1)+
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_", 
           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
           title.col = "Hamilton communities", lwd = 2, 
           legend.col.show = FALSE) + 
  tm_shape(ALL_SCHOOLS_Elem_formapping |> 
             filter(Status != "NoChange" & Status != "Moved")|>
             mutate(Status_change = case_when(Status == "Removed" ~ "Closed",
                                              (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
  tm_symbols(shape = "Status_change",
             size=0.1,
             alpha=0.5,
          title.shape = "Schools that changed status",
          shapes = c(22,21),
          # shapes.labels = c(
          # "New or expanded after 2011",
          # "Closed before 2016"),
          legend.shape.show = FALSE) +
  tmap_defaults +
  tm_layout( bg.color = "grey85",
             panel.label.bg.color = "white",
             panel.label.height = 0.8)

vvim_20112016_CHANGE_plot <-
  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
  tm_polygons(col = "c_v_im_kid_popadj", border.alpha = 0, border.col = "grey20", lwd=0.2,
              palette = "RdBu", midpoint=0,
              breaks = c(-0.1, -0.08, -0.05, 0, 0.1),
              labels = c("[-3.77 to -0.08)", "[-0.08 to -0.05)", "[-0.05 to 0)", 
                         "[0 to 1.46]"),
              title = expression("Change in "*v[i]^m),
              textNA = "No V in 2016 or 2011",
              colorNA = "white")+
  tm_facets(by = c("mode"), nrow=1)+
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_", 
           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
           title.col = "Hamilton communities", lwd = 2) + 
  tm_shape(ALL_SCHOOLS_Elem_formapping |> 
             filter(Status != "NoChange" & Status != "Moved")|>
             mutate(Status_change = case_when(Status == "Removed" ~ "Closed",
                                              (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
  tm_symbols(shape = "Status_change",
             size=0.1,
             alpha=0.5,
          title.shape = "Schools that changed status",
          shapes = c(22,21),
          shapes.labels = c(
          "Closed before 2016",
          "New or expanded after 2011")
  ) +
  tmap_defaults +
  tm_layout( bg.color = "grey85",
             panel.label.bg.color = "white",
             panel.label.height = 0.8)

@@

```{r, eval=FALSE}
tmap_save(Vim_vvim_20112016_CHANGE_plot, 
          paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"),
          height=7,
          width=12, 
          dpi=600)
```

@@

```{r, eval=FALSE}
tmap_save(ttim_GHGim_20112016_CHANGE_plot, 
          paste0(here::here(),"/Figures/ttim_GHGim_20112016_CHANGE_plot.png"),
          height=5.5,
          width=7.5, 
          dpi=600)
```

@@

In this study, we focus on the transportation implications of school closure and consolidation policies. While some research has considered the "steps in reserve", that is, the potential for active travel hidden behind motorized trips [@morencyStepsReserve2009; @morencyHowMany2007], school closures potentially make actual steps vanish, sent to the reserve for the foreseeable future. In exchange for less physical activity, communities get more motorized travel, with the concomitant costs and burdens, including emissions. Quantification of losses in potential active travel and associated growith in carbon emissions is scarce in the literature, and the focus of this paper. In addition to its empirical contributions, the paper makes use of a novel multimodal and singly-constrained accessibility measure, called spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], which we use to estimate the changes in accessibility and travel after a recent wave of school closures/consolidations in Hamilton, Ontario.

The results of the research indicate that, overall, school seats became less spatially available in the city, although this happened unevenly, with some communities being disproportionately affected. Though the school consolidation policy pursued by the province and implemented by the local school board may have resulted in cost-savings in operation and maintenance of facilities, families, students, and communities now pay the price in terms of lost opportunities for healthier active travel, as well as increased emissions from motorized travel.

@@

Hamilton is a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in the province of Ontario, Canada [@governmentofcanadaProfileTableCensus2022]. In 2013, the Hamilton-Wentworth District School Board (HWDSB), the largest public English school board in the city, released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013]. To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of it's elementary schools (i.e., schools attended by students aged 5 to 13) would be subjected to accommodation evaluations, including an "Accommodation Review", the outcome of which historically has been school closures/consolidations [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014]. These decisions pushed by the province were in part motivated by a search for operational savings based on projected reductions in student population and under-utilized school capacity [@craggsAxeCouldFall2012], and Hamilton was particularly affected, as the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014].

Though the Accommodation Review guidelines outline a community-focused process to assess the value that a school provides to students, the community, and the local economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006], in practice residents felt that a wholesome breakdown of costs to repair/maintain schools was not made consistently available [@kleinhuisSeriousConcernsSchool2013]. Further, others pointed to flawed incentive structures motivating the HWDSB: "By closing three schools the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013]. Along with insufficient funding to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015], the operational savings as a case for school closures is a long-standing critique of the school "funding formula" established by the conservative provincial government in 1998 [@mackenzieBlueprintFixOntario2018; @irwinSchoolClosure2012]. With the backdrop of budget cuts, the funding formula pits financing of school maintenance and student needs against the financial compensation of staff and powers of local school boards as they depend on the province for the allocation of resources [@mackenzieBlueprintFixOntario2018].

In the wave of school accommodation reviews between 2013 and 2016, `r ALL_SCHOOLS_Elem %>% filter(Year == "2011" & Level == "Elementary" & Status == "Removed") %>% pull(SchoolID) %>% length()` elementary schools closed in Hamilton. These closures represented a `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` and `r  (( ((avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) - (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum())) / (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum()) ) |> scales::percent()` decline in elementary school locations and capacity between 2011 (before the school closures) and 2016, despite the fact that the overall student population in the city did not decline, and in fact increased slightly between 2011 (`r avail_2011$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students) and 2016 (`r avail_2016$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students).

While accommodation reviews were undertaken to reduce operational costs, how might the closure of schools impact local communities? In this paper, we hypothesize that the school consolidation policy reduced accessibility to schools for many students. By extension, this loss in access and the consequent displacement of active trips to school increased transportation-related emissions. To test these hypotheses, the paper examines the following: how does the (1) spatial availability, (2) loss in walk minutes and (3) associated carbon emissions change for home-to-school trips after a city closes and consolidates schools? Further, this work examines (4) what communities, based on low-income prevalence, experienced the greatest losses as a result of this policy.

Spatial availability, a singly-constrained measure of spatial accessibility [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], is used to quantify accessibility before- and after- school closures, and the associated impacts for the city of Hamilton and the elementary school-aged student population as a result of a policy with narrowly defined benefits. The methods used in this paper can be applied by decision-makers to evaluate the spatial and transport-related impacts of public service provision policy. Furthermore, to increase the diffusion of this analysis, all work is transparent and reproducible, following best practices in the spatial sciences [@brunsdonOpeningPractice2021; @paezOpenSpatial2021c]. Accordingly, all associated code and data will be publicly available within the lead author's [GitHub repository](https://github.com/soukhova)

@@

Our focus is to evaluate the impact of school consolidations between 2011 and 2016 on multimodal school spatial availability and travel-related GHG emission in Hamilton. To facilitate comparison, two scenarios are generated. Each scenario reflects the conditions in 2011 and 2016 respectively, assuming the school configurations in each year, as well as average student population and low-income prevalence of households, residential parcel locations, and estimated modal travel times. The difference between the two scenarios is quantified in terms of accessibility, emissions and equity implications and discussed in context with the school consolidation policy. Data were retrieved at different levels of spatial granularity from the following sources.

@@

@@

\input{TabA1-OTGC-Reg.tex} 
<!--made in the hamONdests repo-->

>>>>

---
title: "School closures and consolidations: the active travel and emission implications of reduced accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
format:
  nature-pdf: 
    journal: "default"
    keep-tex: true
editor: source
execute: 
  freeze: auto

author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
    attributes:
      corresponding: true
  - name: "Christopher D. Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliations:
      - id: 2
        address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"
  - name: "Antonio Páez"
    email: "paezha@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - name: "Moataz Mohamed"
    email: "mmohame@mcmaster.ca"
    affiliations:
      - id: 3
        address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
abstract: | 
  | Reducing emissions and increasing active travel are top priorities for many communities. Often, though, these priorities are confronted with pressures to reduce the fiscal burden of public services, including schools. Public schools have been strained to do more with less, and governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity. The objective of this study is to quantify the travel burdens of school consolidation/closure policies. We do this through the case of Hamilton, a mid-sized city in Ontario, Canada. Between the years 2011 and 2015, 7% of the elementary schools closed in the city. We study how these closures changed the accessibility landscape for the elementary school aged population, as well as the active travel and emission implications. In terms of methods, we use spatial availability, a singly-constrained multimodal measure of spatial accessibility. Thanks to its proportional allocation mechanism, spatial availability allows for an interpretable comparison between 'before' and 'after' the implementation of school closure/consolidation policy. Analysis is conducted at the parcel-level with all school seats being eligible and travel behaviour based on observed home-to-school patterns for motorized and non-motorized modes. We report estimates of system-wide impacts that result from the closures, such as potential growth in GHG emission and reduction of walking minutes. Findings demonstrate that overall, spatial availability decreased in the city. Though the school consolidation policy may have resulted in cost-savings in operation and maintenance of facilities, students, families, and communities pay the price in reduced opportunities for physical activity and greater motorized mobility-related pollution.
keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
---
<!--NOTE WHEN PREPARING THE .TEX for journal submission I ran into errors as their system uses PDFlatex+Bibtex to compile. The following packages only work in lualatex, so do the following:
- remove "  \usepackage[T1]{fontenc}" 
- remove "\usepackage{fontspec}" 
- add "\DeclareUnicodeCharacter{394}{$\Delta$}" to the tex preamble (this is because PHDtex does not recognize unicode? So need to explicitly define) (add to line 5) -->

@@

```{r, include=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggspatial) # Spatial Data Framework for ggplot2
library(here) # A Simpler Way to Find Your Files
library(patchwork) # The Composer of Plots
library(scales) # Scale Functions for Visualization
library(sf) # Simple Features for R
library(stplanr) # Sustainable Transport Planning
library(tidyverse) # Easily Install and Load the 'Tidyverse'
library(tmap) # Thematic Maps
library(units) # Measurement Units for R Vectors
library(miscTools) # Miscellaneous Tools and Utilities
library(huxtable)
library(tiff)
library(grid)
```

@@

```{r descriptive-catchment-schools-plot2, eval=FALSE}
tmap_save(catchment_schools_plot, paste0(here::here(),"/Figures/Fig1.tiff"), height=7, width=12, dpi=600)
```

@@

```{r student-popabs-lowincperc-poprate-plot2, eval=FALSE}
tmap_save(stud_pop_rate_LIM_plot, paste0(here::here(),"/Figures/Fig2.tiff"), height=10, width=12, dpi=500)
```

@@

tmap_save(OD_flow_plot, paste0(here::here(),"/Figures/Fig3.tiff"), height=7, width=12, dpi=600)
```

@@

```{r ploting-empirical-and-theo-curve-TLD-tts, eval=FALSE}
TLD_empirical_wtheo_curve_plot <- ggplot() + 
  geom_histogram(data = od_trips_perc_tt,
                 aes(x = x, 
                     y = ..density..),
                 colour = "grey70",
                 fill = "grey70",
                 boundary = 0,
                 binwidth = 3) +
  geom_vline(data = q_1, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_2, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_3, mapping = aes(xintercept = x), colour = "grey60") +
  geom_line(data = tld_theoretical, aes(x = t, y = f), colour = "blue", size=0.8, alpha=0.6) + 
  xlab("Estimated travel time associated with TTS OD flows (minutes)") +
  ylab("Density") +
  scale_x_continuous(breaks = seq(0,45,10), expand=c(0,0))+
  scale_y_continuous(breaks = seq(0,0.12,0.04), limits = c(0,0.12))+
  facet_grid(~mode~year)+
  geom_text(data = q_1, 
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  geom_text(data = q_2,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  geom_text(data = q_3,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  theme_classic() +
  theme(axis.title = element_text(size = 11),
        panel.border = element_rect(colour = "grey80", fill=NA),
        strip.text = element_text(size=11))
TLD_empirical_wtheo_curve_plot
ggsave(paste0(here::here(),"/Figures/Fig4.tiff"), height=7, width=10, dpi=600)
```
```{r}
ALL_SCHOOLS_Elem_formapping2 <- ALL_SCHOOLS_Elem_formapping |> mutate(Mode = "motorized")
ALL_SCHOOLS_Elem_formapping3 <- ALL_SCHOOLS_Elem_formapping |> mutate(Mode = "non-motorized")
ALL_SCHOOLS_Elem_formapping2 <- rbind(ALL_SCHOOLS_Elem_formapping2, ALL_SCHOOLS_Elem_formapping3)
rm(ALL_SCHOOLS_Elem_formapping3)
```

@@

Vim_vvim_20112016_plot <- tmap_arrange(Vim_20112016_plot, vvim_20112016_plot, nrow=2)

tmap_save(Vim_vvim_20112016_plot, paste0(here::here(),"/Figures/Fig5.tiff"), height=12, width=12, dpi=450)
```

@@

Vim_20112016_CHANGE_plot <-  
  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
  tm_polygons(col = "c_V_im", border.alpha = 0, border.col = "grey20", lwd=0.2,
              palette = "RdBu", #midpoint=0,
              # style="quantile", n=4,
              breaks = c(-20, -4.1, -0.2, 0, 20),
              labels = c("[-208 to -4.1)", "[-4.1 to -0.2)", "[-0.2 to 0)","[0 to 283.6]"),
              title = expression("Change in "*V[i]^m),
              textNA = "No V in 2016 or 2011",
              colorNA = "white")+
  tm_facets(by = c("mode"), nrow=1)+
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_", 
           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
           title.col = "Hamilton communities", lwd = 2, 
           legend.col.show = FALSE) + 
  tm_shape(ALL_SCHOOLS_Elem_formapping |> 
             filter(Status != "NoChange" & Status != "Moved")|>
             mutate(Status_change = case_when(Status == "Removed" ~ "Closed",
                                              (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
  tm_symbols(shape = "Status_change",
             size=0.1,
             alpha=0.5,
          title.shape = "Schools that changed status",
          shapes = c(22,21),
          # shapes.labels = c(
          # "New or expanded after 2011",
          # "Closed before 2016"),
          legend.shape.show = FALSE) +
  tmap_defaults +
  tm_layout(bg.color = "grey85",
             panel.label.bg.color = "white",
            panel.label.height = 0.8)

vvim_20112016_CHANGE_plot <-
  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) + 
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
  tm_polygons(col = "c_v_im_kid_popadj", border.alpha = 0, border.col = "grey20", lwd=0.2,
              palette = "RdBu", midpoint=0,
              breaks = c(-0.1, -0.08, -0.05, 0, 0.1),
              labels = c("[-3.77 to -0.08)", "[-0.08 to -0.05)", "[-0.05 to 0)", 
                         "[0 to 1.46]"),
              title = expression("Change in "*v[i]^m),
              textNA = "No V in 2016 or 2011",
              colorNA = "white")+
  tm_facets(by = c("mode"), nrow=1)+
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_", 
           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
           title.col = "Hamilton communities", lwd = 2) + 
  tm_shape(ALL_SCHOOLS_Elem_formapping |> 
             filter(Status != "NoChange" & Status != "Moved")|>
             mutate(Status_change = case_when(Status == "Removed" ~ "Closed",
                                              (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
  tm_symbols(shape = "Status_change",
             size=0.1,
             alpha=0.5,
          title.shape = "Schools that changed status",
          shapes = c(22,21),
          shapes.labels = c(
          "Closed before 2016",
          "New or expanded after 2011")
  ) +
  tmap_defaults +
  tm_layout(bg.color = "grey85",
             panel.label.bg.color = "white",
            panel.label.height = 0.8)

@@

```{r, eval=FALSE}
tmap_save(Vim_vvim_20112016_CHANGE_plot, 
          paste0(here::here(),"/Figures/Fig6.tiff"),
          height=7,
          width=12, 
          dpi=600)
```

@@

```{r, eval=FALSE}
tmap_save(ttim_GHGim_20112016_CHANGE_plot, 
          paste0(here::here(),"/Figures/Fig7.tiff"),
          height=5.5,
          width=7.5, 
          dpi=600)
```

@@

We focus on the transportation implications of school closure and consolidation policies. While some research has considered the "steps in reserve", that is, the potential for active travel hidden behind motorized trips [@morencyStepsReserve2009; @morencyHowMany2007], school closures potentially make actual steps vanish, sent to the reserve for the foreseeable future. In exchange for less physical activity, communities get more motorized travel, with the concomitant costs and burdens, including emissions. Quantification of losses in potential active travel and associated growth in carbon emissions is scarce in the literature, and the focus of this paper. In addition to its empirical contributions, this work makes use of a novel multimodal and singly-constrained accessibility measure, called spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], which we use to estimate the changes in accessibility and travel after a recent wave of school closures/consolidations in Hamilton, Ontario.

The results of the research indicate that, overall, school seats became less spatially available in the city. This change happened unevenly, with some communities being disproportionately affected. Though the school consolidation policy pursued by the province and implemented by the local school board may have resulted in cost-savings in operation and maintenance of facilities, families, students, and communities now pay the price in terms of lost opportunities for healthier active travel, as well as increased emissions from motorized travel.

@@

Hamilton is a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in the province of Ontario, Canada [@governmentofcanadaProfileTableCensus2022]. In 2013, the Hamilton-Wentworth District School Board (HWDSB), the largest public English school board in the city, released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013]. To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of its elementary schools (attended by students aged 5 to 13) would be subjected to accommodation evaluations, including an "Accommodation Review", the outcome of which historically has been school closures/consolidations [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014]. These decisions, underhandedly supported by the province, were in part motivated by a search for operational savings based on projected reductions in student population and underutilized school capacity [@craggsAxeCouldFall2012]. Hamilton was particularly affected, as the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014].

Though the Accommodation Review guidelines outline a community-focused process to assess the value that a school provides to students, the community, and the local economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006], in practice residents felt that a wholesome breakdown of costs to repair/maintain schools was not made consistently available [@kleinhuisSeriousConcernsSchool2013]. Further, others pointed to flawed incentive structures motivating the HWDSB: "By closing three schools the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013]. Along with insufficient funding to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015], the operational savings as a case for school closures is a long-standing critique of the school "funding formula" established by the conservative provincial government in 1998 [@mackenzieBlueprintFixOntario2018; @irwinSchoolClosure2012]. Within a backdrop of budget cuts, the funding formula pits the financing of school maintenance and student needs against the financial compensation of staff and the powers of local school boards, which depend on the province for the allocation of resources [@mackenzieBlueprintFixOntario2018].

In the wave of school accommodation reviews between 2013 and 2016, `r ALL_SCHOOLS_Elem %>% filter(Year == "2011" & Level == "Elementary" & Status == "Removed") %>% pull(SchoolID) %>% length()` elementary schools closed in Hamilton. These closures represented a `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` and `r  (( ((avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) - (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum())) / (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum()) ) |> scales::percent()` decline in elementary school locations and capacity between 2011 (before the school closures) and 2016, despite the elementary-aged student population in the city slightly increasing between 2011 and 2016( from `r avail_2011$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students to `r avail_2016$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students).

While accommodation reviews were undertaken to reduce operational costs, how might have the closure of schools impacted local communities? In this paper, we hypothesize that the school consolidation policy reduced accessibility to schools for many students. By extension, this loss in access and the consequent displacement of active trips to school increased transportation-related emissions. To test these hypotheses, the paper examines the following: how does the (1) spatial availability, (2) loss in walk minutes and (3) associated carbon emissions change for home-to-school trips after a city closes and consolidates schools? Further, this work examines (4) what communities, based on low-income prevalence, experienced the greatest losses as a result of this policy.

Spatial availability, a singly-constrained measure of spatial accessibility [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], is used to quantify accessibility before- and after- school closures, and the associated impacts for the city of Hamilton and the elementary school-aged student population as a result of a policy with narrowly defined benefits. The methods used in this paper can be applied by decision-makers to evaluate the spatial and transport-related impacts of public service provision policy. Furthermore, to increase the diffusion of this analysis, all work is transparent and reproducible, following best practices in the spatial sciences [@brunsdonOpeningPractice2021; @paezOpenSpatial2021c]. Accordingly, all associated code and data will be publicly available within the lead author's [GitHub repository](https://github.com/soukhova/School-closures-accessibility-impacts)

@@

Our focus is to evaluate the impact of school consolidations between 2011 and 2016 on multimodal school spatial availability and travel-related GHG emission in Hamilton. To facilitate comparison, two scenarios are generated. Each scenario reflects the conditions in 2011 and 2016 respectively, assuming the school configurations in each year, as well as average student population and low-income prevalence of households, residential parcel locations, and estimated modal travel times. The difference between the two scenarios is quantified in terms of accessibility, emissions and equity implications and discussed in context with the school consolidation policy. Data was retrieved at different levels of spatial granularity from the following sources.

@@

\pagebreak

@@

\input{TabA1.tex} 
<!--made in the hamONdests repo-->
